---
title: "dynasymfitr: Fitting dynamically simulated ecological population models to laboratory data"
bibliography: references/dynasymfitr.bib
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Overview

<!-- badges: start -->

<!-- badges: end -->

The goal of `dynasymfitr` is to provide a set of functions to analyze laboratory species interaction experiments in a way that the parameter estimates can be seamlessly integrated into theoretical/computational eco-evo projects. `dynasymfitr` will be developed within the DFG-funded research unit 5726, [DynaSym](https://www.dynasym.uni-konstanz.de/), but is not limited to it.

Estimating density-dependent species interactions is not trivial. The possibly most prominent example is the density-dependent feeding rate, also known as functional response [@Holling1959CharacteristicsSimpleTypes]. Feeding is a temporal process, which requires that the functional response model be modeled as a differential equation describing the decay of the resource over time [@RosenbaumRall2018FittingFunctionalResponses]. While a so-called analytical solution exists for the classical hyperbolic type II functional response [@Royama1971ComparativeStudyModels; @Rogers1972RandomSearchInsect; @Bolker2008EcologicalModelsData], the differential equations of other functional response types need to be solved numerically while being fitted to data [@RosenbaumRall2018FittingFunctionalResponses] to avoid biases in the parameter evaluation. In addition to the above-mentioned problem, if using small organisms such as algae as resources and ciliates as consumers, even in short-term experiments these critters may grow or decline in population size (i.e., the numerical response) due to death and reproduction. This issue would add even more bias to the estimated parameters. However, by simulating the dynamics of resources and consumers explicitly, these biases can be corrected [@RosenbaumRall2018FittingFunctionalResponses]. [DynaSym](https://www.dynasym.uni-konstanz.de/) researchers will predominantly work in aquatic planktonic environments to detect the strength and sign of symbiotic species interactions [@BecksEtAl2025EmergentFeedbackSymbiosis]. Therefore, we will not only add models describing species interactions as mentioned above, but also widely used nutrient uptake models [@Tilman1977ResourceCompetitionPlankton; @HuismanWeissing1999BiodiversityPlanktonSpecies; @Brose2008ComplexFoodWebs; @ZhangBecks2024MechanisticRulesSpecies].

Please note that this package is under development, and we will also add more information to the README in the future.

## License

This package is published under the [**GNU General Public License 3**](https://www.gnu.org/licenses/gpl-3.0.html).

## How to cite

If you use this R-package, please cite it:

Rall, B.C. (2025): dynasymfitr: Fitting dynamically simulated ecological population models to laboratory data. Zenodo. <https://doi.org/10.5281/zenodo.15856656>.

Please cite also the underlying method:

Rosenbaum, B. and Rall, B.C. (2018): Fitting Functional Responses: Direct parameter estimation by simulating differential equations. Methods in Ecology and Evolution 9 (10): 2076–90. <https://doi.org/10.1111/2041-210X.13039>.

### Cite specific versions:

Rall, B.C. (2025): dynasymfitr: Fitting dynamically simulated ecological population models to laboratory data (v0.0.0.9001). Zenodo. <https://doi.org/10.5281/zenodo.15863205>.
Rall, B.C. (2025): dynasymfitr: Fitting dynamically simulated ecological population models to laboratory data (v0.0.0.9000). Zenodo. <https://doi.org/10.5281/zenodo.15856657>.

## Installation

### Install current development version

You can install the latest development version of dynasymfitr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("b-c-r/dynasymfitr@*release") # current release
```

### Install specific versions:

I f you want to install a former version, you may add the specific version number, e.g. like:

``` r
devtools::install_github("b-c-r/dynasymfitr@v0.0.0.9000") # version 0.0.0.9000 (initial release)
```

But see [this thread](https://stackoverflow.com/questions/40179493/r-how-can-i-install-a-specific-release-by-install-github) how to install specific version and even specific commits.

## Examples

### Fitting the generalized functional response

First, load the `dynasymfitr` and the `bbmle` package:

```{r example, message=FALSE, warning=FALSE}
library("dynasymfitr")
# install.packages("bbmle")
library("bbmle")
```

Transfer the data to an object with a shorter name:

```{r}
fr_data <- data_vucic_pestic_et_al_2010_j_anim_ecol
```

Fit the model to the data:

```{r}
fit <- bbmle::mle2(
  minuslogl = calc_gen_fr_nll,
  start = list(
    f_max_log10  = log10(max(fr_data$n_eaten)),
    n_half_log10 = log10(mean(fr_data$n_initial)),
    q = 0.2
  ),
  
  data = list(
    n_eaten = fr_data$n_eaten,
    n_initial = fr_data$n_initial,
    p = rep(1, nrow(fr_data)),
    t_end = rep(1, nrow(fr_data))
  ),
  control = list(reltol = 1e-12, maxit = 1000)
)
```

See the summary table:

```{r}
bbmle::summary(fit)
```

Simulate the best fit:

```{r}
out <- simulate_gen_fr(
 n_initial = 0:1000,                 # vector of initial prey densities
 p = rep(1,1001),                    # fixed predator density
 t_end = rep(1,1001),                # fixed end time of the experiment
 f_max = 10^bbmle::coef(fit)[[1]],   # maximum feeding rate
 n_half = 10^bbmle::coef(fit)[[2]],  # half saturation density
 q = bbmle::coef(fit)[[3]]           # shape parameter (1 = s-shaped)
)
```

Create the regression plot:

```{r fr_example}
# plot results
plot(
 fr_data$n_initial,
 fr_data$n_eaten,
 ylab = "number of prey eaten",
 xlab = "initial prey density"
)
lines(
 out[,1],
 out[,2]
)
```

### Fitting a logistic growth curve assuming pulsed exchange of media

A classical method to design time-series experiments in aquatic systems is to
use chemostats with a constant influx of fresh media and outflux of the media
including the organisms in the experiments. From this outflux, the researcher
can estimate the current density of the organisms.
Alternatively, a fraction of the media is replaced in more or less equally time
intervals. This "pulsed" sampling are comparable to the chemostats, but the
organisms do not have a continuous disturbance, but reoccurring disturbances
events. You analyse such data in the following way.
First, we simulate data:

```{r}
library("bbmle") # for fitting procedure
library("dynasymfitr")

set.seed(667)

data_log_growth <- simulate_logistic_growth_pulsed(
  t_pulse = 0:20,                           # time at which media is exchanged or disturbance happens
  fraction_exchanged = c(0, rep(0.1,20)),   # the fraction of media replaced (or the strength of disturbance)
  n_initial = 1000,                         # the initial resource density
  r = .8,                                   # the intrinsic growth rate
  K = 100000                                # the carrying capacity
)

# create a second time series:
data_log_growth <- rbind(data_log_growth, data_log_growth)
data_log_growth$ts_id <- c(rep("A",21), rep("B",21))
data_log_growth$frac_replaced <- c(0, rep(0.1,20), 0, rep(0.1,20))

# add statistical noise (poisson distributed)
data_log_growth$n_noise <- 10^(rnorm(n = length(data_log_growth$n), mean = log10(data_log_growth$n), sd = .05 ))
```

This example assumes that the time series data has very high densities, e.g.
as often seen with unicells, such as algae, ciliates, or bacteria. The data
can be counts, but also, to keep the values smaller, floating numbers, e.g.
12.3 individuals per microliter. Notably, this approach doesn't allow for
"0" in the density data! Most times, the density data is log-normal distributed,
i.e. low variation if the density is low, high variation in the data if the 
densities are high. We simulate this with the last code line in the code block
above.

Our simulated laboratory data looks quite reasonable:
```{r plot_log_growth_pulsed}
# plot the time series
par(las=1)
plot(
  data_log_growth$t,
  data_log_growth$n_noise,
  ylim = c(0, 125000),
  type = "n",
  pch = 16,
  xlab = "time",
  ylab = "density"
)
points(
  data_log_growth$t[data_log_growth$ts_id == "A"],
  data_log_growth$n_noise[data_log_growth$ts_id == "A"],
  pch = 16,
  col = "blue"
)
points(
  data_log_growth$t[data_log_growth$ts_id == "B"],
  data_log_growth$n_noise[data_log_growth$ts_id == "B"],
  pch = 16,
  col = "red"
)
```

You see two different time series, but the initial density is approximately
the same. In the current version of the package, this requirnment must be 
fulfilled. I aim at making this more flexible later.

### Fit the time series

Like above for the generalized functional response, we use the great `bbmle`
package to fit the data:

``` r
fit_lnorm <- bbmle::mle2(
  minuslogl = calc_logistic_growth_nll_lnorm,
  start = list(
    param_r_log10 = log10(0.8),
    param_K_log10 = log10(100000),
    param_n_initial_log10 = log10(1000),
    param_sd_log10 = log10(0.1)
  ),
  data = list(
    data_time = data_log_growth$t,
    data_density = data_log_growth$n_noise,
    data_frac_exchanged = data_log_growth$frac_replaced,
    data_ts_id = data_log_growth$ts_id
    ),
  fixed = list(
    setting_t_length = 1000
    ),
  control = list(reltol = 1e-12, maxit = 1000)
)
```

```{r message=FALSE, warning=FALSE, include=FALSE, echo = TRUE}
fit_lnorm <- bbmle::mle2(
  minuslogl = calc_logistic_growth_nll_lnorm,
  start = list(
    param_r_log10 = log10(0.8),
    param_K_log10 = log10(100000),
    param_n_initial_log10 = log10(1000),
    param_sd_log10 = log10(0.1)
  ),
  data = list(
    data_time = data_log_growth$t,
    data_density = data_log_growth$n_noise,
    data_frac_exchanged = data_log_growth$frac_replaced,
    data_ts_id = data_log_growth$ts_id
    ),
  fixed = list(
    setting_t_length = 1000
    ),
  control = list(reltol = 1e-12, maxit = 1000)
)
```

When looking at the results, you may note that the values look strange, this is
caused by the fact that the parameters where fitted on a log10 scale, to increase
fitting speed and stability. Moreove it prevents the fitting algorithm to choose
negative values, like e.g. a negative growth rate or carrying capacity.

```{r}
bbmle::summary(fit_lnorm)
```

Back-transforming the parameters reveals that the fit ist quite good:

```{r}
# growth rate simulated value (0.8) is well-matched by fit:
10^bbmle::summary(fit_lnorm)@coef[1]

# carrying capacity simulated value (100000) is well-matched by fit:
10^bbmle::summary(fit_lnorm)@coef[2]

# standard distribution of data (0.05) is well-matched by fit:
10^bbmle::summary(fit_lnorm)@coef[3]

# initial start value of data (1000) is well-matched by fit:
10^bbmle::summary(fit_lnorm)@coef[4]
```

Let us plot the data with the regression line (black line) and a line representing
a simulation in which the researcher would not have disturbed the system (red dashed line):

```{r plot_log_growth_pulsed_fit, echo=FALSE}
par(las=1)
plot(
  data_log_growth$t,
  data_log_growth$n_noise,
  ylim = c(0, 125000),
  type = "n",
  pch = 16,
  xlab = "time",
  ylab = "density"
)
points(
  data_log_growth$t[data_log_growth$ts_id == "A"],
  data_log_growth$n_noise[data_log_growth$ts_id == "A"],
  pch = 16,
  col = "blue"
)
points(
  data_log_growth$t[data_log_growth$ts_id == "B"],
  data_log_growth$n_noise[data_log_growth$ts_id == "B"],
  pch = 16,
  col = "red"
)


## create the regression line
reg_log_growth <- simulate_logistic_growth_pulsed(
  t_pulse = 0:20,                                   # time at which media is exchanged or disturbance happens
  fraction_exchanged = c(0, rep(0.1,20)),           # the fraction of media replaced (or the strength of disturbance)
  n_initial = 10^bbmle::summary(fit_lnorm)@coef[4], # the initial resource density
  r = 10^bbmle::summary(fit_lnorm)@coef[1],         # the intrinsic growth rate
  K = 10^bbmle::summary(fit_lnorm)@coef[2]          # the carrying capacity
)

lines(reg_log_growth$t, reg_log_growth$n, lwd = 2)


## simulate the time series assuming no sampling:
reg_log_growth2 <- simulate_logistic_growth_pulsed(
   t_pulse = seq(0, 20, length = 1000),              # time at which media is exchanged or disturbance happens
   fraction_exchanged = rep(0,1000),                # the fraction of media replaced (or the strength of disturbance)
   n_initial = 10^bbmle::summary(fit_lnorm)@coef[4], # the initial resource density
   r = 10^bbmle::summary(fit_lnorm)@coef[1],         # the intrinsic growth rate
   K = 10^bbmle::summary(fit_lnorm)@coef[2]          # the carrying capacity
 )
lines(reg_log_growth2$t, reg_log_growth2$n, lwd = 2, col = "red", lty = 2)

```

You can find the code either in the according help-file or the README.rmd.


## Funding Information

-   Björn C. Rall gratefully acknowledges the funding by the [**German Science Foundation (DFG) to the Research Unit DynaSym (FOR 5726)**](https://gepris.dfg.de/gepris/projekt/528028597).

## References
